What We Eval in the Shadows

### 2021-04-09 paper/main.tex
tax

The expressions passed to \eval vary widely. In order to categorize them, let us
use a minimization function $min(e)$ which for a given expression $e$ returns a
normal form that abstracts incidental details allowing the reader to focus on
the structure of the evaluated code. The minimization function performs constant
folding of arithmetic and string expressions for base operators, e.g.
$min(\c{1+1})=\c{V}$, value simplification, $min(\c{c(1,2,3+2)})=\c{V}$,
variable absorption, $min(\c{x+y})=\c{X}$, function absorption,
$min(\c{g(f(x),h(z))})=\c{F(F(X))}$ and a number of other simplifications.
Table~\ref{tab:minimizedexpressions} gives the 10 most frequent forms; \#sites
and \%sites are, respectively, the number and ratio of sites receiving arguments
of that form; \#packages is the number of packages with that form; \#operations
is the average number of instructions performed by the interpreter; and \%envir
is the percentage of sites for which the \c{envir} argument is explicitly passed
to \eval and is not \c{parent.frame()}.

\begin{table}[h]\small
\begin{tabular}{|c|r|r|r|r|r|c|}\hline
  $min(e)$& \#sites & \%sites & \#packages & \#operations & \%envir & example\\\hline
%% JAN: simplifying osme mumers and adding %
\c X&\packageMinimizedcallsitesa & 28\% &\packageMinimizedpackagea & 768 &77\% & \c{y+1}\\\hline
\c{F(F(X))} & \packageMinimizedcallsitesb  & 20\% & \packageMinimizedpackageb  & 466K & 55\% & \c{gbov( mean(x), a-1)}\\\hline
\c{V}&\packageMinimizedcallsitesc & 17\% &\packageMinimizedpackagec &\packageMinimizedoperationscRnd & 71\%& \c{c(42,21,0)}\\\hline
\c{F(X)}& \packageMinimizedcallsitesd & 14\% & \packageMinimizedpackaged & \packageMinimizedoperationsdRnd & 75\% & \c{seq\_len(iters)} \\\hline
\c{model.frame} & \packageMinimizedcallsitese & 7\% & \packageMinimizedpackagee & 3K & 21\% & \c{model.frame(formula = Z ~ U)} \\\hline
\c{F()}& \packageMinimizedcallsitesf & 6\% & \packageMinimizedpackagef & \packageMinimizedoperationsfRnd & 73\% & \c{rgamma(3, 2, n = 10L)} \\\hline
\c{\$}& \packageMinimizedcallsitesg & 6\% & \packageMinimizedpackageg & 2K & 78\% & \c{DF\$B} \\\hline
\c{FUN} & \packageMinimizedcallsitesh & 6\% & \packageMinimizedpackageh & \packageMinimizedoperationshRnd & 23\% & \c{function(x, y) x + 3 * y} \\\hline
\c{<-} & \packageMinimizedcallsitesi  & 5\% & \packageMinimizedpackagei & 11K & 73\% & \c{x[1, 2:3, 2:3] <- value}\\\hline
\c{BLOCK} & \packageMinimizedcallsitesj & 4\% & \packageMinimizedpackagej & 440K & 73\% & \c{\{ write.csv(iris, tf) \textbackslash n file.size(tf) \}} \\\hline
%% \c X&\packageMinimizedcallsitesa &\packageMinimizedpropsitesa\% &\packageMinimizedpackagea &\packageMinimizedoperationsaRnd &\packageMinimizedpercentenvirnonparenta\% & \c{y+1}\\\hline
%% \c{F(F(X))} & \packageMinimizedcallsitesb  & \packageMinimizedpropsitesb & \packageMinimizedpackageb  & \packageMinimizedoperationsbRnd & \packageMinimizedpercentenvirnonparentb & \c{gbov( mean(x), a-1)}\\\hline
%% \c{V}&\packageMinimizedcallsitesc &\packageMinimizedpropsitesc &\packageMinimizedpackagec &\packageMinimizedoperationscRnd &\packageMinimizedpercentenvirnonparentc& \c{c(42,21,0)}\\\hline
%% \c{F(X)}& \packageMinimizedcallsitesd & \packageMinimizedpropsitesd & \packageMinimizedpackaged & \packageMinimizedoperationsdRnd & \packageMinimizedpercentenvirnonparentd & \c{seq\_len(iters)} \\\hline
%% \c{model.frame} & \packageMinimizedcallsitese & \packageMinimizedpropsitese & \packageMinimizedpackagee & \packageMinimizedoperationseRnd & \packageMinimizedpercentenvirnonparente & \c{model.frame(formula = Z ~ U)} \\\hline
%% \c{F()}& \packageMinimizedcallsitesf & \packageMinimizedpropsitesf & \packageMinimizedpackagef & \packageMinimizedoperationsfRnd & \packageMinimizedpercentenvirnonparentf & \c{rgamma(3, 2, n = 10L)} \\\hline
%% \c{\$}& \packageMinimizedcallsitesg & \packageMinimizedpropsitesg & \packageMinimizedpackageg & \packageMinimizedoperationsgRnd & \packageMinimizedpercentenvirnonparentg & \c{DF\$B} \\\hline
%% \c{FUN} & \packageMinimizedcallsitesh & \packageMinimizedpropsitesh & \packageMinimizedpackageh & \packageMinimizedoperationshRnd & \packageMinimizedpercentenvirnonparenth & \c{function(x, y) x + 3 * y} \\\hline
%% \c{<-} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedoperationsiRnd & \packageMinimizedpercentenvirnonparenti & \c{x[1, 2:3, 2:3] <- value}\\\hline
%% \c{\{\}} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedoperationsjRnd & \packageMinimizedpercentenvirnonparentj & \c{\{ write.csv(iris, tf) \textbackslash n file.size(tf) \}} \\\hline
\end{tabular}
\caption{Minimized expressions} \label{tab:minimizedexpressions}
\end{table}

15 normal forms. We detail the interesting cases next.

@@

\medskip\noindent\EE{$min(e)=\c{V}$} Expression that represent values are
frequently passed to \eval. The majority of those, \packageValOneNodePercent,
are inline constants (integers or doubles). The rest trivially evaluate to a
value.\footnote{True as long as base functions such as \c{+} are not redefined.
They typically aren't, but this is a limitation of this categorization.} In our
corpus, \packageNbCallSitesUniqueActualValue call sites only ever see a simple
values; the median number of times they are exercised is
\packageMedianRunSitesUniqueActualValue, suggesting a possible lack of coverage.
However, manual inspection reveals cases such as:
\begin{lstlisting}
 f1 <- eval(paste("A~",paste(paste(names(X[,-1])),collapse="+")))
\end{lstlisting}
  This concatenates to a string and \eval simply returns it as is.
  From twenty randomly selected \evals, only \packageUsefulValueEvalPercent of
  them needed to use \eval.

\medskip\noindent\EE{$min(e)=\c{X}$} Variables are the most common form. They
subsume values, thus \c{X} includes arithmetic expressions in which a variable
occurs. \packageNbSymbolVarSitePercent of variable expression are symbols, \ie
just an \c{x}. As a variable may refer to a promise, it may trigger computation.

\medskip\noindent\EE{$min(e)=\c{\$}$} A name lookup, \c{x\$f}, can be used to
access variable in an environment or a column in a data frame, this also include
vector indexing operators, \c{[]} and \c{[[]]}. A name lookup subsumes
variables.

\medskip\noindent\EE{$min(e)=$~\c{<-}} Assignments account for
\packageAssignSitesPercent of \eval sites, they are \c{<-}, {\tt <<-} and
\c{assign}, and they subsume named lookups.

\medskip\noindent\EE{$min(e)=\c{F()}$} A plain function call which may have some
values as argument but does not include variables. It should be noted that
function name require variable lookups -- we distinguish those as they usually
do not trigger computation.

\medskip\noindent\EE{$min(e)=\c{F(X)}$} Function calls whose arguments may include
variable references.

>>>>

The expressions passed to \eval vary widely. In order to categorize them, let us
use a minimization function $min(e)$ which for a given expression $e$ returns a
normal form that abstracts incidental details allowing the reader to focus on
the structure of the evaluated code. The minimization function performs constant
folding of arithmetic and string expressions for base operators, e.g.
$min(\c{1+1})=\c{V}$, value simplification, $min(\c{c(1,2,3+2)})=\c{V}$,
variable absorption, $min(\c{x+y})=\c{X}$, function absorption,
$min(\c{g(f(x),h(z))})=\c{F(F(X))}$ and a number of other simplifications.
Table~\ref{tab:minimizedexpressions} gives the 10 most frequent forms; \#sites
and \%sites are, respectively, the number and ratio of sites receiving arguments
of that form; \#packages is the number of packages with that form; \#operations
is the average number of instructions performed by the interpreter; and \%envir
is the percentage of sites for which the \c{envir} argument is explicitly passed
to \eval and is not \c{parent.frame()}. The example column shows one sample
expression $e$ that is normalized to the particular form.

\begin{table}[h]\small
\begin{tabular}{|c|r|r|r|r|r|c|}\hline
  $min(e)$& \#sites & \%sites & \#packages & \#operations & \%envir & example\\\hline
%% JAN: simplifying osme mumers and adding %
\c X&\packageMinimizedcallsitesa & 28\% &\packageMinimizedpackagea & 768 &77\% & \c{y+1}\\\hline
\c{F(F(X))} & \packageMinimizedcallsitesb  & 20\% & \packageMinimizedpackageb  & 466K & 55\% & \c{gbov( mean(x), a-1)}\\\hline
\c{V}&\packageMinimizedcallsitesc & 17\% &\packageMinimizedpackagec &\packageMinimizedoperationscRnd & 71\%& \c{c(42,21,0)}\\\hline
\c{F(X)}& \packageMinimizedcallsitesd & 14\% & \packageMinimizedpackaged & \packageMinimizedoperationsdRnd & 75\% & \c{seq\_len(iters)} \\\hline
\c{model.frame} & \packageMinimizedcallsitese & 7\% & \packageMinimizedpackagee & 3K & 21\% & \c{model.frame(formula = Z ~ U)} \\\hline
\c{F()}& \packageMinimizedcallsitesf & 6\% & \packageMinimizedpackagef & \packageMinimizedoperationsfRnd & 73\% & \c{rgamma(3, 2, n = 10L)} \\\hline
\c{\$}& \packageMinimizedcallsitesg & 6\% & \packageMinimizedpackageg & 2K & 78\% & \c{DF\$B} \\\hline
\c{FUN} & \packageMinimizedcallsitesh & 6\% & \packageMinimizedpackageh & \packageMinimizedoperationshRnd & 23\% & \c{function(x, y) x + 3 * y} \\\hline
\c{<-} & \packageMinimizedcallsitesi  & 5\% & \packageMinimizedpackagei & 11K & 73\% & \c{x[1, 2:3, 2:3] <- value}\\\hline
\c{BLOCK} & \packageMinimizedcallsitesj & 4\% & \packageMinimizedpackagej & 440K & 73\% & \c{\{ write.csv(iris, tf); file.size(tf) \}} \\\hline
%% \c X&\packageMinimizedcallsitesa &\packageMinimizedpropsitesa\% &\packageMinimizedpackagea &\packageMinimizedoperationsaRnd &\packageMinimizedpercentenvirnonparenta\% & \c{y+1}\\\hline
%% \c{F(F(X))} & \packageMinimizedcallsitesb  & \packageMinimizedpropsitesb & \packageMinimizedpackageb  & \packageMinimizedoperationsbRnd & \packageMinimizedpercentenvirnonparentb & \c{gbov( mean(x), a-1)}\\\hline
%% \c{V}&\packageMinimizedcallsitesc &\packageMinimizedpropsitesc &\packageMinimizedpackagec &\packageMinimizedoperationscRnd &\packageMinimizedpercentenvirnonparentc& \c{c(42,21,0)}\\\hline
%% \c{F(X)}& \packageMinimizedcallsitesd & \packageMinimizedpropsitesd & \packageMinimizedpackaged & \packageMinimizedoperationsdRnd & \packageMinimizedpercentenvirnonparentd & \c{seq\_len(iters)} \\\hline
%% \c{model.frame} & \packageMinimizedcallsitese & \packageMinimizedpropsitese & \packageMinimizedpackagee & \packageMinimizedoperationseRnd & \packageMinimizedpercentenvirnonparente & \c{model.frame(formula = Z ~ U)} \\\hline
%% \c{F()}& \packageMinimizedcallsitesf & \packageMinimizedpropsitesf & \packageMinimizedpackagef & \packageMinimizedoperationsfRnd & \packageMinimizedpercentenvirnonparentf & \c{rgamma(3, 2, n = 10L)} \\\hline
%% \c{\$}& \packageMinimizedcallsitesg & \packageMinimizedpropsitesg & \packageMinimizedpackageg & \packageMinimizedoperationsgRnd & \packageMinimizedpercentenvirnonparentg & \c{DF\$B} \\\hline
%% \c{FUN} & \packageMinimizedcallsitesh & \packageMinimizedpropsitesh & \packageMinimizedpackageh & \packageMinimizedoperationshRnd & \packageMinimizedpercentenvirnonparenth & \c{function(x, y) x + 3 * y} \\\hline
%% \c{<-} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedoperationsiRnd & \packageMinimizedpercentenvirnonparenti & \c{x[1, 2:3, 2:3] <- value}\\\hline
%% \c{\{\}} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedoperationsjRnd & \packageMinimizedpercentenvirnonparentj & \c{\{ write.csv(iris, tf) \textbackslash n file.size(tf) \}} \\\hline
\end{tabular}
\caption{Minimized expressions} \label{tab:minimizedexpressions}
\end{table}

\noindent
We detail these forms and discuss their implication for the behavior of \eval.

@@

\medskip\noindent\EE{$min(e)=\c{V}$} Expression that represent values are
frequently passed to \eval -- they represent 17\% of the calls. The majority of
those, \packageValOneNodePercent, are inline constants (integers or doubles).
The rest trivially evaluate to a value, \eg~\c{1+1}.\footnote{True as long as
base functions such as \c{+} are not redefined. They typically aren't, but this
is a limitation of this categorization.} In our corpus,
\packageNbCallSitesUniqueActualValue call sites only ever see a simple value.
Manual inspection reveals cases such as the following
\begin{lstlisting}
 f1 <- eval(paste("A~",paste(paste(names(X[,-1])),collapse="+")))
\end{lstlisting}
where the argument to \eval is a string which \eval simply returns. From twenty
randomly selected \evals, \packageUsefulValueEvalPercent of them do not need
\eval. For the other cases, either we are dealing with a value that needs to be
constructed dynamically or the value is a default case that sometimes is
replaced by a more interesting expression.

\medskip\noindent\EE{$min(e)=\c{X}$} Variables lookups are the most common form,
they are found in 28\% of the sites. This form includes simple variable reads,
\eg~\c{x}, those are \packageNbSymbolVarSitePercent of \c X. The form also
subsumes \c{V}, so it includes a mixture of arithmetic expressions,
\eg~\c{x+y+1}. The operations allowed are limited to built-in arithmetics. It is
noteworthy that, while most \c{X}\!s evaluate in a single step, the variable can
be bound to a promise, and accessing it may trigger evaluation of that promise,
thus resulting in an arbitrary amount of computation.

\newpage
\medskip\noindent\EE{$min(e)=\c{\$}$} This form extends \c X to include lookup
with the dollar operator, \eg~\c{x\$f}, and vector indexing, \eg~\c{x[42]} or
\c{x[[24]]}. As with \c X we allow arithmetics and values in this form. Lookup
occurs in 6\% of the sites. Arbitrary computations can occur if one of the
variables is bound to a promise.

\medskip\noindent\EE{$min(e)=$~\c{<-}} This form includes both assignments
operator, the direct assignment {\tt <-}, and assignment to the parent
environment {\tt <\,\!<-}, and the \c{\$} form. Assignments occur in 4\% of the
sites. They represent the most obvious source of side-effects.

\medskip\noindent\EE{$min(e)=\c{F()}$} This form captures simple function calls,
\eg~\c{f(2)}, with neither variables or assignments. Or, more specifically it
allows for variables in the function position but not in arguments. Usually,
looking up function names does not trigger computation, but that is not a given
if the function is returned by a promise or if the function name is shadowed by
a promise containing a value, then computation will occur. This form occurs in
6\% of sites.

\medskip\noindent\EE{$min(e)=\c{F(X)}$} This form allows function calls whose
arguments may include variable references. Assignments are excluded from this
form. It occurs in 14\% of the sites in the corpus.
